%************************************************
\chapter{General discussion and conclusions}\label{ch:discussion}
%************************************************

Life history theory seeks to explain how life history traits are selected in
concert in the face of constraints, trade-offs.
The theory has evolved from simple explanations based on single traits such as
clutch size \citep{Lack1946,Skutch1949} to broader explanations based on the
allometric relations among life history traits and body size
\citep{Western1979}, r-K selection where traits are selected according to the
population density \citep{Pianka1970}, age-specific mortality
\citep{Gadgil1970,Stearns1976,Charlesworth1980} and more recently, adding
behaviour as a factor interacting with the evolution of the life history traits
\citep{Ricklefs2002,Reale2010a,Sol2016}.
Despite the progress made during all these years, there is a gap in the theory
regarding the effects of the life history on the response of organisms to
environmental changes.
My thesis contributes to the advance of this field in two ways.
First, by defining a demographically meaningful axes of life history variation
in birds and confirming the existence of trade-offs among traits and
restrictions for the existing combinations of life history traits (Chapter
\ref{ch:LHaxes}).
Second, by understanding the mechanisms by which life history interacts with
behaviour and clarifying how together they affect how species cope with new
environmental conditions (Chapters \ref{ch:LH-Behaviour model} and
\ref{ch:POLS}).

\bigskip

The first objective of this thesis was to describe the diversity of life
history in birds. Despite that it is not the first attempt to do so
\citep{Saether1987,Gaillard1989,Saether2000,Jeschke2009}, my work in chapter
\ref{ch:LHaxes} represents a significant improvement over previous works in
three important ways.
First, I used newly available statistical methods that control for the
phylogenetic structure of the data by means of phylogenetic principal component
analysis \citep{Revell2009a} and phylogenetic least square regressions
\citep{Ho2014}.
Second, the number of species in the compiled dataset is one order of magnitude
larger than previous works. Finally, the availability of demographic data and
the link with different combinations of life history traits allowed to
objectively define the axes of life history variation that better describe
demographic features of the species instead of using traits according to their
availability.
In addition of improving the quantification of the already known fast-slow
continuum
\citep{Stearns1983a,Saether1987,Gaillard1989,Oli2004,Dobson2007,Jeschke2009},
my analyses also allowed to describe additional axes of life history variation
such as the iteroparity axis \citep{Gaillard1989}, an axis related to the
trade-off between offspring quantity and quality
\citep{Promislow1990,Bielby2007,Dobson2007}, and finally an axis that reflects
the lifelong productivity of the species in terms of offspring number and egg
mass produced relative to body size.
The availability of a large data base of the species position in each life
history axis opens new possibilities to understand the implications of the life
history in an evolutionary and ecological framework.
Life history in birds is not as diverse as in other groups, perhaps due to the
fly constraints \citep{Gaillard1989,Healy2014}.
Future works could apply a similar methodology to other groups such as mammals
for which traits and demographic data is already available
\citep{Myhrvold2015,Salguero-Gomez2016}.

\bigskip

A second objective of the thesis was to explore how life history affects the
response of species to environmental changes. Previous attempts to elucidate
the question showed contradictory patterns about the effects of the fast-slow
axis on establishment success of introduced species (non significant for birds
\citep{Blackburn2009a,Sol2012a} but significant for mammals and reptiles
\citep{Capellini2015,Allen2017}), or the ability to colonize urban habitats
(non significant for birds \citep{Sol2014} but significant for mammals
\citep{Santini2019}).
As I argued in chapters \ref{ch:LH-Behaviour model} and \ref{ch:POLS}, one
possible explanation is the omission of an important factor that affects and
is affected by the species position in the fast-slow axis: Behaviour.

As a first approach to the question, I explored mechanisms by which life history
and behaviour can interact to influence demography in novel environments
by means of a theoretical individual based model simulating the introduction of
species with different life histories in a new environment (Chapter
\ref{ch:LH-Behaviour model}). The model shows that under maladaptive scenarios,
where the match of the phenotype to the environment is insufficient, it pays
off to have a slow life history that increases the value of adults over the
value of offspring even at the cost of decreasing reproduction.
This is in part owing to the demographic consequences of the life-history
strategy itself and in part owing to the higher benefits of behavioural
responses for slow species in comparison to fast species.
The notion that slow animals exposed to novel environments generally gain
greater benefits from behavioural responses has been suggested in previous
studies (reviewed in \citet{Sol2016}). Animals at the ‘slow’ extreme of the
fast–slow continuum are generally believed to explore more accurately the
environment and exhibit better performance in learning than those at the ‘fast’
extreme. One reason for this difference being that they can develop
disproportionally larger brains, which has been shown to enhance the capacity
to innovate and learn
\citep{Lefebvre1997,Reader2002,Overington2009,Reader2011}. The model in chapter
\ref{ch:LH-Behaviour model} supports the idea that life history and behaviour
are not independent and that studied together can help better understand the
mechanisms by which they affect the responses to environmental changes, an issue
of great relevance in the context of human induced rapid environmental changes
where behavioural responses can be the key to adapt to novel conditions. Future
studies can use similar approaches to look for demographic and ecological
consequences of other relevant axes of life history variation such as
iteroparity and offspring quality-quantity as described in Chapter
\ref{ch:LHaxes}. In particular, iteroparity seems a relevant axis to persist in
novel environments as comparative studies have shown for invasive species
\citep{Sol2012a} an urban dwellers \citep{Sol2014,Sayol2020}.

A second approach to understand how life history and behaviour affect the
response to environmental changes was developed in Chapter \ref{ch:POLS}. In
this chapter we used a comparative analysis of flight initiation distances for
birds in rural and urban habitats. The results show the existence of a
peace-of-life syndrome (POLS) predicted by theory where slow-lived species tend
to be more risk-averse than fast-lived species \citep{Reale2010a}.
Furthermore, the POLS structure vanishes in urbanized environments due to
slow-lived species adjusting their flight distances based on the perception of
risk. Even though there is no evidence in birds that slow species are better
urban colonisers \citep{Sol2014}, the fact that slow species have a more plastic
behaviour supports the idea that slow species can potentially better adapt to
environmental changes.
On the other hand, the pattern seems to revert in mammals, where species with
traits related to the fast end of the fast-slow continuum are better urban
dwellers in some groups \citep{Santini2019}.
The contradictory results about whether fast or slow strategies are better to
deal with environmental changes is still open. The fact that there is no clear
pattern in birds could be because, even though there is a fast-slow continuum
among species, the birds as a group are mostly in the slow end compared to
mammals \citep{Healy2014}.

\bigskip

Human activities are altering natural environments through a wide range of
processes, from land use changes to introduction of exotic species and climate
change. Each type of changes can affect differently the resource availability
and the age-specific mortality. Thus, the effects on the species and their
responses should be different. Thus, there is no single life history strategy to
deal with all changes. Rather, the nature of the changes will determine if the
environment will become more or less predictable, including novel resource
opportunities or qualitatively different threats such as new predators to which
species are not adapted.
In unpredictable environments, bet-hedging strategies can be a good strategy
\citep{Starrfelt2012}. Other less well-studied strategies for unpredictable
environments include plastic life histories mediated by behaviour such as the
ability to module the reproductive effort by skipping reproduction in bad years,
the so called ``storage effect'' \citep{Forcada2008}.
In environments with high pressure from new predators such as cats, juvenile
mortality by predators can be compensated by increasing reproduction if enough
resources are available \citep{Yeh2004} displacing the species towards the fast
end of the fast-slow continuum. Also, the increase of the juvenile survival
will have less impact for slow species which exhibit a life history buffer
against the effects of demographic stochasticity and are
less sensitive to changes in juvenile survival as I showed in Chapter
\ref{ch:LH-Behaviour model}.
Species can respond to predictable changes through adjustments on the life
history \citep{Evans2005} or through behavioural plasticity mechanisms such as
learning \citep{Laundre2001,Evans2012}.
In the case of rapid changes for which there is not enough time for evolutive
responses, behavioural plastic responses may be the only way to adapt to the
new opportunities or threats \citep{Sol2009}, and there is a growing number of
evidences that behaviour affects and is affected by life history
\citep{Sol2016}.
Thus, when we examine how life history affects the responses to environmental
changes, we are considering not only life history mechanisms but also mechanisms
related to behavioural responses.

Probably, the answer to the question of which life history strategies are
better to respond to environment changes is context-dependent, and only by
carefully thinking about the relevant mechanisms, including behavioural
responses, for each scenario of environmental change and the taxonomic group of
study, we can refine the understanding of the effects of life history on the
ability of the species to survive to environmental changes.


\clearpage

\subpdfbookmark{Conclusions}{Conclusions}
\section*{Conclusions}

\begin{itemize}

  \item \textbf{Chapter 2:}
  \begin{itemize}
    \item Not all combinations of life history traits exist in nature. The
variation of life history traits is organized in different axes caused by
trade-off and phylogenetic constraints.
    \item One of the main axes of traits' covariation is the fast-slow,
described recurrently since 1983 by a varying set of traits more often
justified by the traits data availability than for ecological reasons. This
axis is related to the survival-fecundity trade-off.
Defining the fast-slow  axis in a demographically meaningful way (i.e.
optimizing the correlation of the resulting axis with the elasticity to adult
survival or generation time) allowed me to build an objective and more
ecologically relevant characterisation of the axis.
    \item The remaining variation in life history traits once the fast-slow
is ruled out is organized in three other axes of variation, sorted by variance
explained:

    \begin{itemize}
      \item Iteroparity, describes the degree of concentration of the
reproductive effort in few or many breeding attempts.
      \item Lifelong potential productivity, related to the lifelong investment
in reproduction in terms of the number of offspring and the egg mass production
relative to the body mass.
      \item Offspring quality-quantity trade-off, which sorts species along a
continuum with species with a large relative egg mass and small clutch size in
one end, and species with large clutch size and small relative egg size at the
other end.
    \end{itemize}
  \end{itemize}


  \item \textbf{Chapter 3:}
  \begin{itemize}
    \item Theoretical models help to investigate the effects and mechanisms
that affect the demography of species in novel or unfamiliar environments.
    \item Under maladaptive scenarios where the mismatch of phenotype to the
environment is insufficient, simulations suggest that slow lived species,
for which adult have more value than offspring, have more chances to be
established.
    \item Behavioural responses interact with life history to influence the
persistence of populations in novel and unknown environments. The benefits of
learning behaviours are greater for slow strategies. Behaviours such
as skipping a reproductive event, can improve the probabilities to be
established for slow species while being detrimental for fast species. And
finally, innate responses in a context of novel environments can be beneficial
or impact negatively the probabilities to establish a population if preferences
do not match habitat quality (ecological trap), being fast strategies more
impacted by ecological traps on scenarios with higher offspring mortality and
slow strategies more impacted on scenarios with higher adult mortality.
  \end{itemize}


  \item \textbf{Chapter 4:}
  \begin{itemize}
    \item  Slow-lived species tend to have a more risk-averse behaviour than
fast-lived species.
    \item The relationship between flight initiation distance (FID) and the
fast-slow continuum is largely mediated by differences in body size among
species. Possible causes include a higher likelihood to be detected by
predators, lower maneuverability to escape when attacked and higher energetic
costs associated with flight.
    \item Flight initiation distance is shorter in urban than in rural
environments. While slow-lived species showed shorter or larger FID
according to the perception of risk, fast-lived species did not accommodate
their FID to the degree of human frequentation. The changes in FID observed in
slow-lived species may reflect plastic adjustments, selection, and/or a
non-random sorting of individuals by behaviours that affect invasion success.
  \end{itemize}


%   \item \textbf{Chapter 5:}
%   \begin{itemize}
%     \item ...
%     \item ...
%   \end{itemize}

\end{itemize}
